---
title: "Case Study - Adverse Events vs Country"
author: "Fred"
date: "8/21/2020"
output:
  html_document: default
  pdf_document: default
---
## R Packages used in this analysis
```{r, message=FALSE, warning=FALSE}
library(httr)
library(jsonlite)
library(knitr)
library(kableExtra)
library(dplyr)
```

# Problem Statement
The database of FDA adverse event report systems (FAERS) contain a list of the patient reactions and the country where a report comes from. The research question of this study is, are different adverse events reported in different countries? In other words, the association between adverse event types and countries is to be tested.

# Methods
The variables to be tested are categorical variables. In statistics, contingency tables are used to summarize the relationship between categorical variables, and a chi-square test can then be conducted on contingency tables to test wheter or not an association exists between variables. When the evidence of association is found, pairwise two sample proportion tests can be used to compare different proportion of a specific adverse event type between any given two countries.

The data used in this study is the drug event endpoint and ranges from the frist quarter of 2004 to the second quarter of 2020.

# Results
## Data Exploration: Countries Reporting to the FAERS
```{r, warning=FALSE}
# A function to query FAERS database via openFDA API 
GetResponse = function(URL) {
  response = GET(URL)
  # Request failed if status code != 200
  if(response$status_code != 200)
    return(NULL)
  response = content(response, as = "text")
  response = fromJSON(response)
  response = response$results
  return(response)
}

Country_table = GetResponse("https://api.fda.gov/drug/event.json?api_key=ts0BKaDMfrjs2adslwvDJWIM8fGcngovS1QX5jD4&count=primarysource.reportercountry.exact&limit=1000")
kable(Country_table, format = "html", row.names = FALSE) %>% kable_styling() %>% scroll_box(width = "75%", height = "250px")
top10countries = c("US", "GB", "CA", "JP", "FR", "DE", "IT", "BR", "ES", "NL")
```

```{R}
grep_coutries = c('US|(UNITED STATES)', 'GB|(UNITED KINGDOM)', 'CA|CANADA', 'JP|JAPAN', 'FR|FRANCE', 'DE|GERMANY', 'IT|ITALY', 'BR|BRAZIL', 'ES|SPAIN', 'NL|NETHERLANDS')
sum(Country_table$count[grep(paste(grep_coutries, collapse = '|'), Country_table$term)])/sum(Country_table$count)
```
From the query results, we can see that some countries have multiple names, e.g., "US" and "United States", "United Kingdom" and "GB", "JP" and "Japan", etc, which is to be combined in the data preprocessing step.  

Although there are 447 different country names were retrieved, most of the reports were submitted by the top several countries. The top 10 countries (United States, United Kingdom, Canada, Japan, France, Germany, Italy, Brazil, Spain, and Netherlands) account for 89% of all adverse events reported. To simplify the analysis, I focused on the top 10 countries.

## Data Exploration: Patient Reactions Reporting to the FAERS
```{R}
AE_table = GetResponse("https://api.fda.gov/drug/event.json?api_key=ts0BKaDMfrjs2adslwvDJWIM8fGcngovS1QX5jD4&count=patient.reaction.reactionmeddrapt.exact&limit=1000")
kable(AE_table, format = "html", row.names = FALSE) %>% kable_styling() %>% scroll_box(width = "75%", height = "250px")
```

The count queries returns up to 1000 results, and I cannot paginate through "count" queries with "skip" parameter. So only top 1000 adverse events can be counted. The approach in this study and the codes below can analyze arbituary number of adverse events, however, for simplicity, only the top 10 adverse events were chosen for analysis.

## Contingency Table Creation
```{R}
TOP_N_AE = 10 # top N adverse events to be analyzed. Can be set to any arbituary natural number.

# Create a contigency table of country x AE
mAEbyCountry = matrix(rep(NA, 10*(1+TOP_N_AE)), ncol=10)
row.names(mAEbyCountry) = c(AE_table$term[1:10], 'OTHERS')
colnames(mAEbyCountry) = top10countries
queries_coutries = c('US+"UNITED+STATES"', 'GB+"UNITED+KINGDOM"', 'CA+CANADA', 'JP+JAPAN', 'FR+FRANCE', 'DE+GERMANY', 'IT+ITALY', 'BR+BRAZIL', 'ES+SPAIN', 'NL+NETHERLANDS')

## Count top N adverse events for each country
for (j in 1:10) {
  for (i in 1:TOP_N_AE) {
    query = paste("https://api.fda.gov/drug/event.json?api_key=ts0BKaDMfrjs2adslwvDJWIM8fGcngovS1QX5jD4&search=(primarysource.reportercountry:", queries_coutries[j], ')+AND+(patient.reaction.reactionmeddrapt:"', gsub(' ', '+', AE_table$term[i]),'")&count=patient.reaction.reactionmeddrapt.exact&limit=1', sep='')
    result = GetResponse(query)
    if(!is.null(result)) {
      mAEbyCountry[i,j] = result$count[1]
    }
  }
}

## Count "OTHERS" adverse events for each country
for(j in 1:10) {
  mAEbyCountry[(TOP_N_AE+1),j] = sum(Country_table$count[grep(grep_coutries[j], Country_table$term)]) - sum(mAEbyCountry[1:TOP_N_AE,j])
}
```

The table below is the contingency table which displays frequency distribution of the variables (countries and adverse events).  

```{R}
kable(mAEbyCountry, format = "html", caption="Contingency Table", row.names = TRUE) %>% kable_styling() %>% scroll_box(width = "100%", height = "450px")
```

We can also visualize the distribution of the adverse events in each of the countries. From the pie charts below, we can see different patterns. In Canada, the top 10 adverse events account for more than two thirds of all reported events, while in Italy and Spain, the same adverse events account for only a small proportion (less than 5%).

```{R, out.width='100%', fig.asp=0.5, fig.ncol = 1, fig.cap="output", fig.align = "center"}
par(mfrow = c(2, 5))
for (j in 1:10) {
 pie(mAEbyCountry[,j], radius=1, cex=0.5, main=paste("AE in", top10countries[j])) 
}
```


## Statistical Tests

A chi-square test of the contingency table indicates that there is significant association between adverse events coutries (p-value<2.2e-16).
```{R}
# chi-squared test of association: countries vs adverse events
chisq.test(mAEbyCountry)
```

Furthermore, we can compare the difference in distribution of a specific adverse event by pairwise proportion test. Take death for example. From the p-value table, we can see except for Italy and Spain, both of which have very low death events (0.11%), all other pairs of comparisons are significant. P-values shown below are adjusted by Bonferroni correction. 

```{R}
# pairwise proportion test of "death" adverse event
ls.test = pairwise.prop.test(mAEbyCountry["DEATH",], apply(mAEbyCountry, 2, sum), "bonferroni")
kable(ls.test$p.value, format = "html", caption="p-values of pairwise proportion test", row.names = TRUE) %>% kable_styling() %>% scroll_box(width = "100%", height = "400px")
```

The bar plots display the distribution of the death events in each country. We can see that Netherlands have significantly higher death adverse events rate (15%) than other countries.

```{R}
barplot(mAEbyCountry["DEATH",]/apply(mAEbyCountry, 2, sum)*100, main = "Proportion of Death Adverse Events in Different Countries", ylab="Proportion (%)")
```


# Discussion
There are some limitations in this study. First, the FAERS data is not fully validated. In the exploration of the variable of country, we have seen there are several similar names for the same country. Although I have combined some of them in the data preprocessing step, yet a complete thesaurus (including misspelled names) is needed to avoid any omission. The variable of adverse events have the same issue. Also the adverse events are underreported. If the missing data mechanism is not completely random, then our analysis may be biased. E.g., the very low death event rates in Italy and Spain might be caused by underreporting and not at random. Third, for simplicity, I only chose the top 10 adverse events to analyze. It happens that the significant difference was detected. If not, since the majority of adverse events were under the category of "others", I have to include more types of adverse events to accept or reject the null hypothesis.

# Conclusion
In this case study, a contingency table and chi-square test are used to test the hypothesis about the association between adverse events and countries. According to the results, different adverse events are reported in different countries (p-value<2.2e-16). 

